---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

1. Load the data (i.e. `read.csv()`)

```{r loading}

file <- read.csv("activity.csv", header=TRUE, colClasses = c("numeric", "Date",     "numeric"))

```
2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r process}

data <- file[!is.na(file$steps), ]

```
## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day and 

```{r mean}

sum.total.days <- tapply(data$steps, data$date, sum)

hist(sum.total.days, main="Histogram mean total number of steps taken per day", xlab= "Mean total number of steps taken per day")

```

2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r mean.report}

mean(sum.total.days)
median(sum.total.days)

```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r daily.activity}

mean.intervals <- tapply(data$steps, data$interval, mean)

plot(unique(data$interval), mean.intervals, type = "l", main = "Average Daily Activity", xlab = "Intervals", ylab = "Average across all days")

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r daily.activity.max}

names(mean.intervals[mean.intervals == max(mean.intervals)])

```
## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)

``` {r calculate.misssing.values}

length(is.na(file$steps))

```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

``` {r filling.values}

mean.intervals.df <- aggregate(list(data$steps), list(data$interval), mean)

colnames(mean.intervals.df) <- c("interval", "mean.total")

data.na <- file[is.na(file$steps), ]

for (i in unique(file[is.na(file$steps), ]$interval)){
  data.na[data.na$interval == i, 1] <- mean.intervals.df[mean.intervals.df$interval == i, 2]
  }

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

``` {r df.final}

data.final <- rbind(data, data.na)
data.final <- data.final[order(data.final$date, data.final$interval), ]

summary(data.final)
head(data.final, 20)

```

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

``` {r weekdays}

data.final <- mutate(data.final, weekday = weekdays(date))
data.final[data.final$weekday == "Saturday" | data.final$weekday == "Sunday", "weekday.factor"] <- c("WE")
data.final[!(data.final$weekday == "Saturday" | data.final$weekday == "Sunday"), "weekday.factor"] <- c("WD")

str(data.final)
head(data.final, 20)

```

2. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using **simulated data**:

``` {r plot}

weekdays <- subset(data.final, data.final$weekday.factor == "WD")
weekends <- subset(data.final, data.final$weekday.factor == "WE")

weekdays.mean <- tapply(weekdays$steps, weekdays$interval, mean)
weekends.mean <- tapply(weekends$steps, weekends$interval, mean)

par(mfrow=c(2, 1), bty = "l")
plot(unique(weekdays$interval), weekdays.mean, type = "l", main = "Average Daily Activity on Regular Weekdays", xlab = "Intervals", ylab = "Average across all days")
plot(unique(weekends$interval), weekends.mean, type = "l", main = "Average Daily Activity on Weekends", xlab = "Intervals", ylab = "Average across all days")


```